Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-09-06 poster/poster.tex
before i realize it's portrait
poster good first version

\usetheme{RJH}
\usepackage[utf8]{inputenc}
\usepackage[frenchb]{babel}
\usepackage[orientation=portrait,size=a2,scale=1.2]{beamerposter}
\usepackage[absolute,overlay]{textpos}
\usepackage{url}

\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage[labelformat=empty]{caption}
\usepackage{framed}
\usepackage{wrapfig}

@@

@@

%% Column 1 ==================================================================

\begin{textblock}{0.32}(0.01,0.14)

%% Signal processing ---------------------------------------------------------

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions. \textbf{We derive a
three-level decomposition of the  information jointly provided by all input
variables about the output in terms of
\begin{itemize}
\item[] {\color{lightgreen}i) the MDI importance of each input variable,}
\item[] {\color{lightblue}ii) the degree of interaction of an input variable with the other input variables,}
\item[] {\color{lightred}iii) the different interaction terms of a given degree.}
\end{itemize}}
We then
show that this MDI importance of a variable is equal to zero if and only if the
variable is irrelevant and that the MDI importance of a relevant variable is
invariant with respect to the removal or the addition of irrelevant variables.
We illustrate these properties on a simple example and discuss how they may
change in the case of  non-totally randomized trees such as Random Forests and
Extra-Trees.

\vspace{2.5cm}

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions. \textbf{We derive a
three-level decomposition of the  information jointly provided by all input
variables about the output in terms of
\begin{itemize}
\item[] {\color{lightgreen}i) the MDI importance of each input variable,}
\item[] {\color{lightblue}ii) the degree of interaction of an input variable with the other input variables,}
\item[] {\color{lightred}iii) the different interaction terms of a given degree.}
\end{itemize}}
We then
show that this MDI importance of a variable is equal to zero if and only if the
variable is irrelevant and that the MDI importance of a relevant variable is
invariant with respect to the removal or the addition of irrelevant variables.
We illustrate these properties on a simple example and discuss how they may
change in the case of  non-totally randomized trees such as Random Forests and
Extra-Trees.

@@

%% Col 1 Row 2

\begin{textblock}{0.32}(0.01,0.42)

%% Col 1 Row 3

\begin{textblock}{0.32}(0.01,0.71)

\begin{center}
\includegraphics[width=0.75\linewidth]{images/network.png}
\end{center}

%% Column 2 ==================================================================

\begin{textblock}{0.32}(0.34,0.14)

In this work, we propose \textbf{a simple yet effective solution} to the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
\textbf{two steps:
\begin{itemize}
\item[] {\color{lightgreen}i) processing the raw signals to detect neural peak activities,}
\item[] {\color{lightblue}ii) inferring the degree of association between neurons from partial
correlation statistics.}
\end{itemize}}
We summarises the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method, and
finally compares our results with respect to other inference methods.

\begin{textblock}{0.32}(0.34,0.31)

%% Introduction ------------------------------------------------------------

@@

%% Col 2 Row 2

\begin{textblock}{0.32}(0.34,0.525)

%% Connectome inference ------------------------------------------------------

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions. \textbf{We derive a
three-level decomposition of the  information jointly provided by all input
variables about the output in terms of
\begin{itemize}
\item[] {\color{lightgreen}i) the MDI importance of each input variable,}
\item[] {\color{lightblue}ii) the degree of interaction of an input variable with the other input variables,}
\item[] {\color{lightred}iii) the different interaction terms of a given degree.}
\end{itemize}}
We then
show that this MDI importance of a variable is equal to zero if and only if the
variable is irrelevant and that the MDI importance of a relevant variable is
invariant with respect to the removal or the addition of irrelevant variables.
We illustrate these properties on a simple example and discuss how they may
change in the case of  non-totally randomized trees such as Random Forests and
Extra-Trees.

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions.

@@

%% Col 2 Row 3

\begin{textblock}{0.32}(0.34,0.71)

%% Column 3 ==================================================================

\begin{textblock}{0.32}(0.67,0.14)

%% Experiments ---------------------------------------------------------------

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions. \textbf{We derive a
three-level decomposition of the  information jointly provided by all input
variables about the output in terms of
\begin{itemize}
\item[] {\color{lightgreen}i) the MDI importance of each input variable,}
\item[] {\color{lightblue}ii) the degree of interaction of an input variable with the other input variables,}
\item[] {\color{lightred}iii) the different interaction terms of a given degree.}
\end{itemize}}
We then
show that this MDI importance of a variable is equal to zero if and only if the
variable is irrelevant and that the MDI importance of a relevant variable is
invariant with respect to the removal or the addition of irrelevant variables.
We illustrate these properties on a simple example and discuss how they may
change in the case of  non-totally randomized trees such as Random Forests and
Extra-Trees.

\vspace{2.5cm}

Despite growing interest and practical use in various scientific areas, \textbf{variable
importances derived from tree-based ensemble methods are not well understood
from a theoretical point of view}. In this work we characterize the Mean Decrease
Impurity (MDI) variable importances as measured by an ensemble of totally
randomized trees in asymptotic sample and ensemble size conditions. \textbf{We derive a
three-level decomposition of the  information jointly provided by all input
variables about the output in terms of
\begin{itemize}
\item[] {\color{lightgreen}i) the MDI importance of each input variable,}
\item[] {\color{lightblue}ii) the degree of interaction of an input variable with the other input variables,}
\item[] {\color{lightred}iii) the different interaction terms of a given degree.}
\end{itemize}}
We then
show that this MDI importance of a variable is equal to zero if and only if the
variable is irrelevant and that the MDI importance of a relevant variable is
invariant with respect to the removal or the addition of irrelevant variables.
We illustrate these properties on a simple example and discuss how they may
change in the case of  non-totally randomized trees such as Random Forests and
Extra-Trees.

@@

%% Col 3 Row 2

\begin{textblock}{0.32}(0.67,0.42)

%% Col 3 Row 3

\begin{textblock}{0.32}(0.67,0.71)

\begin{block}{Conclusion \phantom{p}}

Its simplified variant outperforms other
network inference methods while its optimized version proved to be the best method
on the Connectomics Challenge. Given its simplicity and good performance, we
therefore believe that the methodology presented in this work
would constitute a solid and easily-reproducible baseline for further work in
the field of connectome inference.\\[2ex]

\textbf{Acknowledgments}\\
A. Joly and G. Louppe are research fellows of the FNRS, Belgium.  A. Sutera is a
recipient of an FRIA fellowship of FRS-FNRS, Belgium. This work is supported by
PASCAL2 and the IUAP DYSCO, initiated by the Belgian State, Science Policy
Office.

@@

>>>>

\usetheme{RJH}
\usepackage[utf8]{inputenc}
\usepackage[frenchb]{babel}
\usepackage[orientation=portrait,size=a2,scale=1.2]{beamerposter}
\usepackage[absolute,overlay]{textpos}
\usepackage{url}
\usepackage[T1]{fontenc}

\usepackage{bbold}

\usepackage{wrapfig, blindtext}
\usepackage{ragged2e}

\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage[labelformat=empty]{caption}
\usepackage{framed}
%\usepackage{wrapfig}

@@

\newenvironment{shaded2}{%
  \def\FrameCommand{\fboxsep=\FrameSep \colorbox{blue!20}}%
  \MakeFramed {\FrameRestore}}%
 {\endMakeFramed}

@@

\begin{textblock}{0.35}(0.01,0.115)

\vspace{-0.4cm}
\begin{center}

\begin{minipage}{0.95\linewidth}
\begin{shaded}
In this work, we propose \textbf{a simple yet effective solution} to the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
\textbf{two steps:
\begin{itemize}
\item[] {\color{lightgreen}i) processing the raw signals to detect neural peak activities,}
\item[] {\color{lightblue}ii) inferring the degree of association between neurons from partial
correlation statistics.}
\end{itemize}}
We summarises the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method, and
finally compares our results with respect to other inference methods.

\end{shaded}
\end{minipage}
\end{center}

@@

%% Context ------------------------------------------------------------

\begin{textblock}{0.62}(0.37,0.115)

@@

%% Signal processing -----------------------------------------------------

\begin{textblock}{0.485}(0.01,0.26)

Under the simplifying assumption that neurons are on-off units, characterised
by short periods of intense activity, or peaks, and longer periods of
inactivity, the first part of our algorithm consists of cleaning the raw
fluorescence data.
More specifically, time-series are processed using standard
signal processing filters in order to : 
\textbf{
\begin{itemize}
\item[] {\color{lightgreen}i) remove noise mainly due to fluctuations independent of calcium, calcium fluctuations independent of spiking activity, calcium fluctuations in nearby tissues that have been mistakenly captured, or simply by the imaging process ;}
\item[] {\color{lightblue}ii) to account for fluorescence low decay ; and}
\item[] {\color{lightred}iii) to reduce the importance of high global activity in the network.}
\end{itemize}}
\vspace{-2pt}

\begin{textblock}{0.485}(0.01,0.395)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\begin{center}
\includegraphics[width=0.85\linewidth]{images/original_curve.pdf}
\end{center}
\end{shaded}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.25,0.395)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\vspace{15pt}
\hspace{1cm} {\small \textbf{Original curve}}
\vspace{30pt}
\end{shaded}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.25,0.43)
\begin{minipage}{0.48\linewidth}
\begin{shaded2}
\begin{center}

\includegraphics[width=0.85\linewidth]{images/filter_curve.pdf}
\end{center}
\vspace{15pt}
\end{shaded2}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.01,0.495)
\begin{minipage}{0.48\linewidth}
\begin{shaded2}
\vspace{1pt}
{\color{lightgreen} \textbf{Applying a low-pass filter}}
\vspace{-5pt}
\begin{align*}
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i,\\
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\end{align*}
\vspace{-10pt}
\end{shaded2}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.01,0.54)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\begin{center}

\includegraphics[width=0.85\linewidth]{images/diff_curve.pdf}
\end{center}
\end{shaded}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.25,0.54)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\vspace{3pt}
{\color{lightgreen} \textbf{Applying a high-pass filter}}
\begin{align*}
g(x^{t}_{i}) = x^{t}_i - x^{t-1}_i,
\end{align*}
\vspace{5pt}
\end{shaded}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.25,0.58)
\begin{minipage}{0.48\linewidth}
\begin{shaded2}
\begin{center}

\includegraphics[width=0.85\linewidth]{images/threshold_curve.pdf}
\end{center}

\end{shaded2}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.01,0.64)
\begin{minipage}{0.48\linewidth}
\begin{shaded2}
\vspace{7pt}
{\color{lightblue} \textbf{Applying a hard-threshold filter}}
\vspace{-5pt}
\begin{align*}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align*}
\vspace{8pt}
\end{shaded2}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.01,0.685)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\begin{center}
\includegraphics[width=0.85\linewidth]{images/original_curve.pdf}
\end{center}
\vspace{-14pt}
\end{shaded}
\end{minipage}
\end{textblock}

\begin{textblock}{0.485}(0.25,0.685)
\begin{minipage}{0.48\linewidth}
\begin{shaded}
\vspace{15pt}
{\color{lightred} \textbf{Magnify the importance of spikes that occur in cases of low global activity}}
\begin{align*}
w(x^{t}_i) = (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align*}\vspace{42pt}
\end{shaded}
\end{minipage}
\end{textblock}

@@

% \begin{columns}
% \begin{column}{0.01\linewidth}

% \end{column}
% \begin{column}{0.5\linewidth}
% \includegraphics[width=0.9\linewidth]{images/original_curve.pdf}\\

% \begin{align*}
% f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i,\\
% f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
% \end{align*}

% \includegraphics[width=0.9\linewidth]{images/diff_curve.pdf}\\[2ex]
% \begin{align*}
% h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
% \end{align*}
% \includegraphics[width=0.9\linewidth]{images/weights_curve.pdf}\\

% \end{column}

% \begin{column}{0.5\linewidth}

% Original signal \\

% \includegraphics[width=0.9\linewidth]{images/filter_curve.pdf}\\[2ex]

% \begin{align*}
% g(x^{t}_{i}) = x^{t}_i - x^{t-1}_i,
% \end{align*}

% \includegraphics[width=0.9\linewidth]{images/threshold_curve.pdf}\\[2ex]

% \begin{align*}
% w(x^{t}_i) = (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
% \end{align*}

% \end{column}
% \end{columns}

%% Connectome inference --------------------------------------------------

\begin{textblock}{0.485}(0.505,0.26)

\includegraphics[width=\linewidth]{images/partial2.png}

Given the assumption of random variables drawn from the same time-invariant joint probability distribution, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
\textit{Partial correlation} coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:

\begin{align*}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{align*}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$.\\

Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, i.e., $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$.

\begin{itemize}
\item[{\color{green} \cmark}] Only direct associations
\item[{\color{green} \cmark}] Filter out spurious indirect effects
\item[{\color{green} \cmark}] Symmetric (i.e. $p_{i,j}=p_{j,i}$)\\[5ex]
\item[{\color{red} \xmark}] Edge orientation
\item[{\color{red} \xmark}] Sensitive to the value of parameters in the filtering process
\end{itemize}

@@

%% Experiment ------------------------------------------------------------

\begin{textblock}{0.69}(0.01,0.78)

\vspace{-5pt}
\begin{wrapfigure}{r}{0.7\linewidth}
\centering
\tiny
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{8}{c|}{Performance with partial correlation}\\ \hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
with \textit{Method} $\backslash$ on \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
No  filtering       					& 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
$ h \circ g \circ f_1$                  & 0.923 & 0.925 & 0.923 & 0.922 & 0.311 & 0.315 & 0.313 & 0.304\\
$ w \circ h \circ g \circ f_1$          & 0.931 & 0.929 & 0.928 & 0.926 & 0.326 & 0.323 & 0.319 & 0.303\\
+ PCA         							& 0.932 & 0.930 & 0.928 & 0.926 & 0.355 & 0.353 & 0.350 & 0.333\\
Averaging           					& 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  0.390 &  0.385 & 0.375\\
Full method           					& \textbf{0.943} & \textbf{0.942} & \textbf{0.942} & \textbf{0.939} & \textbf{0.403} & \textbf{0.404} & \textbf{0.398} & \textbf{0.388}\\
\hline
PC & 0.886 & 0.884 & 0.891 &  0.877 & 0.153 & 0.145 & 0.170 & 0.132\\
GTE & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142\\
GENIE3 & 0.892 & 0.891 & 0.887 & 0.887 & 0.232 & 0.221 & 0.237 & 0.215 \\
\hline
\end{tabular}
\end{wrapfigure}

These results clearly show the importance of the filters and PCA (especially for AUPRC). Taking the average over
various parameter settings gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row (``Full method'') shows the
final performance of the method specifically tuned for the challenge
(see paper for all details). Although this
tuning was decisive to obtain the best performance in the challenge,
it does not significantly improve either AUROC or AUPRC.

Partial correlation and averaged partial correlation clearly outperform all other methods on all datasets. The
improvement is more important in terms of AUPRC than in terms of AUROC. As
expected, Pearson correlation performs very poorly in terms of AUPRC. GTE and
GENIE3 work much better, but these two methods are nevertheless clearly below
partial correlation.

@@

\begin{textblock}{0.28}(0.71,0.78)

\begin{block}{Conclusion \phantom{p}}
\vspace{-0.4cm}
\begin{center}

\begin{minipage}{0.95\linewidth}
\begin{shaded}

Its simplified variant outperforms other
network inference methods while its optimized version proved to be the best method
on the Connectomics Challenge. Given its simplicity and good performance, we
therefore believe that the methodology presented in this work
would constitute a solid and easily-reproducible baseline for further work in
the field of connectome inference.
\end{shaded}
\end{minipage}
\end{center}

\vspace{-0.3cm}

\begin{textblock}{1}(0.01,0.99)
{\fontsize{12}{10}\textbf{Acknowledgments:}}
{\fontsize{8}{10}\textbf{A. Joly and G. Louppe are research fellows of the FNRS, Belgium. A. Sutera is a
recipient of an FRIA fellowship of FRS-FNRS, Belgium. This work is supported by
PASCAL2 and the IUAP DYSCO, initiated by the Belgian State, Science Policy
Office.}}
\end{textblock}
